Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-07 supplement.tex
add Tim's section to conclusion
label
fix abstract and margins

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{xspace}

@@

$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
archaic admixture \citep{Ragsdale2019-nt}, and this statistic is closely
related to $S^*$ statistics used to scan for introgressed haplotypes
\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
widely used in demographic inference involving ancient DNA and many samples, as
$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
\citep{Lipson2020-cq}. Therefore, the set of statistics used here encompass
multiple features of genetic data that have been used to infer models of
archaic admixture and population structure involving many populations.

@@

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \citep{1000_Genomes_Project_Consortium2015-zq}.
This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of modern humans and Neanderthal
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for archaic introgression from a population whose split from modern
humans is at least as old as that of the human-Neanderthal split
\citet{Durvasula2020-td,Yang2012-ze}.
Sites with no calls in the Vindija Neanderthal were excluded from
this analysis.

@@

For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of modern humans, with no structure in the stem and no
archaic admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of modern humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
Southern and West/Eastern African populations.

To include population structure in early human history, we considered multiple
parameterizations of models that allowed more than a single stem population. In
general, stem populations were allowed to vary in their sizes, split times, and
migration rates, with parameterizations flexible enough to encompass proposed
scenarios of either archaic admixture or population structure, both connected
by gene flow or with isolation between stems.

@@

In another parameterization of early structure, a secondary stem population
(stem 2) contributed ancestry to present-day populations via a series of
instantaneous admixture events (i.e., pulse admixture or “merger” events) to
lineages leading to sampled present-day African populations. Merger events were
allowed to occur in one or more of the Nama, Mende, and Gumuz branches, as well
as the branch of East and West Africans prior to their split. Those admixture
events were allowed to occur at any time along those branches, and with any
proportions, and stem 2 was allowed to split from the primary stem at any time
before subsequent divergences (and either before or after the split of the
Neanderthal branch). We tested models that allowed migration between the early
stems, before subsequent splits and admixture events, as well as models that
were restricted to isolation between stem branches. Depending on the specific
parameters, such models encompass commonly-considered ghost archaic admixture
scenarios (e.g., if a long-isolated lineage more recently contributes a
minority of ancestry to one or more populations), as well as relatively simple
fragmentation-coalescence scenarios.

@@

A deep branch is assigned to a contemporary population if at least one sample
from that population has ancestry that passes through that branch.
\citet{Speidel2019-nj} show that deep branches with lower ends more recent than
the Neanderthal introgression event are enriched for Neanderthal-matching
alleles in Eurasian populations, while a large majority of deep branches in
1000 Genomes West African populations do not match either the Neanderthal or
Denisovan samples. This observation was taken as further evidence for deep
population structure or archaic admixture in West African populations from an
unidentified hominin unrelated to the Neanderthal/Denisovan complex.

@@

\subsection{What does this mean for expectation of morphology btw branches?}

\aprcomment{Tim to fill in, and edit section header}

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization. These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic admixture from an unidentified hominin species
within Africa.

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{xspace}

@@

$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
archaic hominin admixture \citep{Ragsdale2019-nt}, and this statistic is closely
related to $S^*$ statistics used to scan for introgressed haplotypes
\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
widely used in demographic inference involving ancient DNA and many samples, as
$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
\citep{Lipson2020-cq}. Therefore, the set of statistics used here encompass
multiple features of genetic data that have been used to infer models of
archaic hominin admixture and population structure involving many populations.

@@

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \citep{1000_Genomes_Project_Consortium2015-zq}.
This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of humans and Neanderthals
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for introgression from an archaic hominin population that split from humans at least as far back in time as the human-Neanderthal split
\citet{Durvasula2020-td,Yang2012-ze}.
Sites with no calls in the Vindija Neanderthal were excluded from
this analysis.

@@

For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
hominin admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of modern humans, with no structure in the stem and no
archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of modern humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
Southern and West/Eastern African populations.

To include population structure in early human history, we considered multiple
parameterizations of models that allowed more than a single stem population. In
general, stem populations were allowed to vary in their sizes, split times, and
migration rates, with parameterizations flexible enough to encompass proposed
scenarios of either archaic hominin admixture or population structure, both connected
by gene flow or with isolation between stems.

@@

In another parameterization of early structure, a secondary stem population
(stem 2) contributed ancestry to present-day populations via a series of
instantaneous admixture events (i.e., pulse admixture or “merger” events) to
lineages leading to sampled present-day African populations. Merger events were
allowed to occur in one or more of the Nama, Mende, and Gumuz branches, as well
as the branch of East and West Africans prior to their split. Those admixture
events were allowed to occur at any time along those branches, and with any
proportions, and stem 2 was allowed to split from the primary stem at any time
before subsequent divergences (and either before or after the split of the
Neanderthal branch). We tested models that allowed migration between the early
stems, before subsequent splits and admixture events, as well as models that
were restricted to isolation between stem branches. Depending on the specific
parameters, such models encompass commonly-considered scenarios for admixture from a ghost archaic hominin population (e.g., if a long-isolated lineage more recently contributes a
minority of ancestry to one or more populations), as well as relatively simple
fragmentation-coalescence scenarios.

@@

A deep branch is assigned to a contemporary population if at least one sample
from that population has ancestry that passes through that branch.
\citet{Speidel2019-nj} show that deep branches with lower ends more recent than
the Neanderthal introgression event are enriched for Neanderthal-matching
alleles in Eurasian populations, while a large majority of deep branches in
1000 Genomes West African populations do not match either the Neanderthal or
Denisovan samples. This observation was taken as further evidence for deep
population structure or archaic hominin admixture in West African populations from an
unidentified hominin unrelated to the Neanderthal/Denisovan complex.

@@

\subsection{What does this mean for expectation of morphology btw branches?}
\label{morpho_diff}
\aprcomment{Tim to fill in, and edit section header}

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization. These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic hominin admixture from an unidentified hominin species
within Africa.
